---
title: "PA1_template"
author: "Kartik Deshmukh"
date: "March 6, 2016"
output: html_document
---
### Reproducible Research Assignment 1

#### Mean and Median of Total number of steps taken per day

```{r Daily mean and median, echo=TRUE}
# Reading activity data
activity <- read.csv("activity.csv", 
                      header = TRUE, sep = ",", na.strings = "NA")

# Converting into date and time format
act_date <- as.Date(activity$date)

daily_steps <- sapply(split(activity,activity$date), 
                    function(x) x <- {return (sum(x$steps, na.rm = TRUE))})

#Plotting Histogram
hist(daily_steps, col = "red", 
     ylab = "Frequency", xlab = "Daily Steps", breaks = 25, main = "Total Steps Distribution")

daily_mean <- as.integer(round(mean(daily_steps), 2))
daily_median <- median(daily_steps)
```
The mean of total number of steps taken per day is **`r daily_mean`**.  
The median of total number of steps taken per day is **`r daily_median`**.

#### Average Daily Activity Pattern
```{r Average Daily Activity pattern,echo=TRUE}
avg_step_int <- sapply(split(activity,activity$interval), 
                    function(x) x <- {return (mean(x$steps, na.rm = TRUE))})

intervals <- unique(activity$interval)
plot(intervals, avg_step_int, type = "l",
     ylab = "Avg. Steps", 
     xlab = "Intervals")
max_interval <- intervals[match(max(avg_step_int), avg_step_int)]
```
The 5 minute interval containing maximum number of steps is **`r max_interval`**.

#### Inputting missing values and the effect on dataset.
```{r Missing Values,echo=TRUE}
no_values <- sum((is.na(activity$steps) == TRUE))

for (i in which(is.na(activity$steps))) {
  activity[i,1] <- avg_step_int[match(activity[i,3], intervals)]
}

# Revised total daily steps
daily_steps <- sapply(split(activity,activity$date), 
                    function(x) x <- {return (sum(x$steps, na.rm = TRUE))})

#Plotting Histogram
hist(daily_steps, col = "red", 
     ylab = "Frequency", xlab = "Daily Steps", breaks = 25, main = "Total Steps Distribution")

daily_mean <- as.integer(round(mean(daily_steps), digits = 3))
daily_median <- as.integer(round(median(daily_steps), digits = 3))
```
Total rows of missing values are **`r no_values`**.  

The new mean of total number of steps taken per day is **`r daily_mean`**.    

The new median of total number of steps taken per day is **`r daily_median`**.  

These new values differ from their earlier estimates after filling missing data values in data set with the means for their particular intervals.  

Also, the revised histogram has smoothed out due to filling the missing values.

#### Weekdays and Weekend Activity difference
```{r weekday and weekend activity}
library(lattice)
fac_days <- factor ((weekdays(act_date, abbreviate = FALSE) %in% c('Monday',   'Tuesday', 'Wednesday', 'Thursday', 'Friday')), 
                    levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))

activity_new <- cbind(activity, fac_days)

# Splitting data by weekdays and weekends
act_weeks <- split(activity_new, activity_new$fac_days)

#Calculating mean average for intervals in weekdays
avg_step_int_wday <- sapply(split(act_weeks$weekday, act_weeks$weekday$interval), 
                       function(x) x <- {return (list( mean = mean(x$steps, na.rm = TRUE),
                                                       interval = x[1,3], week = x[1,4]))})

#Calculating mean average for intervals in weekends
avg_step_int_wend <- sapply(split(act_weeks$weekend, act_weeks$weekend$interval), 
                       function(x) x <- {return (list(mean = mean(x$steps, na.rm = TRUE),
                                                      interval = x[1,3], week = x[1,4]))})
# Inverting columns and rows of data frame
avg_step_int_wday <- as.data.frame(t(avg_step_int_wday))
avg_step_int_wend <- as.data.frame(t(avg_step_int_wend))

act_final <- rbind(avg_step_int_wend, avg_step_int_wday)

week_type <- factor(act_final$week, levels = c(1,2), labels = c("Weekdays", "Weekends"))

xyplot(act_final$mean~act_final$interval|week_type,xlab = "interval", ylab = "No. of Steps",
       layout = c(1,2), panel = function(x, y) { panel.loess(x, y) })
```

